Main Analysis for Biracial Grade Retention Project

### 2022-09-08 analysis/analysis_main.Rmd
Update explanation of example figure

```{r}
marg(model4, "race", type="effects",
     at_var_interest=c("White","Black/White","Black"))
marg(model4, "race", type="effects",
     at_var_interest=c("Black","Black/Latino","Latino"))
marg(model4, "race", type="effects",
     at_var_interest=c("Asian","White/Asian"))
marg(model4, "race", type="effects",
     at_var_interest=c("Indigenous","White/Indigenous"))
marg(model4, "race", type="effects",
     at_var_interest=c("White","White/Indigenous"))
marg(model4, "race", type="effects",
     at_var_interest=c("White","White/Latino"))
marg(model4, "race", type="effects",
     at_var_interest=c("Latino","White/Latino"))
```

@@

I can do this by calculating the *average adjusted prediction* of the dependent variable across all of the possible race groups. The average adjusted prediction (this is the stata-ese language - it is also called variously a predictive margin among other things). The adjusted predition in general is just the predicted value of the dependent variable for some set of values on the covariates. In this case, I want to predict on the level of the response to get predicted probabilities of being held back. The average adjusted prediction essentially gets the adjusted prediction for every observation in the data based on their covariates and then takes the average. Essentially, this is equivalent to treating every respondent as if they belonged to group A when calculating the average adjusted prediction for group A. The average adjusted prediction is directly related to the *average marginal effect* (AME) in that the AME for a group difference is simply the difference in the average adjusted prediction across groups.

@@

In order to calculate all the values that I need, I have created a function that will put this into a nice table for later plotting.

@@

Now I can use this nice table to plot out the results. I illustrate what this basic plot looks like below for the case of black/white individuals.

@@

This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions (although the SEs are slightly higher due to survey design effects of clustering and variable weights). The grey bands for the monoracial groups bookmark the range of results we might expect and the red band gives what would be the midpoint value between the groups, adjusting for statistical uncertainty. I can then see where the actual biracial estimate falls within this range. The results here indicate that biracial black/white students have probabilities of grade retention much closer to the lower values of white than black students. However, their probability is still a bit higher than that for whites.

Why I am using 83% (technically 83.4%) confidence intervals? This is a nice visual trick. People often incorrectly believe that overlapping 95% confidence intervals on two means indicates that the difference in the means is not statistically significant at $p<.05$. This is not always true and leads in practice to a much lower $\alpha$ for rejection. A much better approximation is given when one uses1.386 rather than 1.96 in estimating the confidence interval, which produces an 83.4% confidence interval. The intuition behind this technique is given nicely [here](https://chris-said.io/2014/12/01/independent-t-tests-and-the-83-confidence-interval-a-useful-trick-for-eyeballing-your-data/). Because I use this interval, we can know that if intervals (or bands) do not overlap, then the difference in means between the two groups is statistically significant at $p<.05$. In this case, the biracial estimate is statistically significantly different from both monoracial estimates as well as the "halfway" estimate. In short, biracial black/white students are held back at rates much closer but somewhat higher than whites, rather than blacks.

Its important to note that this technique is just an approximation because it depends on the assumption that the standard errors for the two means are the same, which is not going to be true here. However, it should give results that are quite close. Correcting for unequal SEs is far more complicated because the interval used will vary for any pairwise comparison which I am not doing here.

@@

```{r cond-means-model4, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Estimates are based on models that adjust for state, year, current grade, metropolitan status, gender, foreign born status, parent's foreign born status, English proficieny, parent's English proficiency, parent's education, family income, home ownership, and family structure."}
coef_table_mod4 <- calculate_marg_means(model4)

ggplot(coef_table_mod4, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table_mod4, multiracial))+
  geom_point()+
  coord_flip()+
  facet_wrap(~mrace, ncol=3, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average adjusted predicted probability of being behind expected grade")
```

@@

```{r cond-means-model-compare, fig.height=12, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Each panel is based on a different model that accounts for certain variables. Subsequent panels cumulatively include all prior terms."}
#calculate remaining marginal means
coef_table_mod1 <- calculate_marg_means(model1)
coef_table_mod2 <- calculate_marg_means(model2)

@@

* Lower than both groups
* Same as lower group
* Lower half
* Halfway between
* Upper half
* Same as higher
* Higher than both groups

@@

|                   | Lower than both | Same as lower | Lower half | Halfway | Upper half | Same as higher | Higher than both |
|-------------------|:---------------:|:-------------:|:----------:|:-------:|:----------:|:--------------:|:----------------:|
| Black/White       |        X*       |               |            |         |            |                |                  |
| White/Asian       |        X*       |       X       |            |         |            |                |                  |
| White/Latino      |        X        |       X*      |            |         |            |                |                  |
| Black/Latino      |        X        |       X*      |      X     |    X    |            |                |                  |
| Indigenous/Asian  |        X        |       X*      |      X     |         |            |                |                  |
| Indigenous/White  |                 |               |      X*    |         |            |                |                  |
| Indigenous/Latino |                 |               |      X     |    X*   |      X     |                |                  |
| Black Asian       |                 |       X       |      X     |    X*   |      X     |       X        |                  |
| Black/Indigenous  |                 |       X       |      X     |    X*   |      X     |       X        |                  |
| Latino/Asian      |                 |       X       |      X     |    X    |      X     |       X        |        x*        |

@@

* For White/Black students, grade retention is clearly in the lower-half between Whites and the midway point rather than below whites for high school students. Notably, this is also where the white/black grows the largest. So the advantage that these students have relative to everyone else only holds for elementary and middle school. Nonetheless, their outcomes are still much closer to White students than Black students.
* For White/Indigenous students, the placement varies across panels. In elementary school, these students are indistinguishable from White students. They are perfectly in the middle in middle school, and then in the lower half in high school. So it seems that their lowest outcomes happen earlier and on average you end up with a "lower half" across all grades.

>>>>

@@

I can do this by calculating the *average adjusted prediction* of the dependent variable across all of the possible race groups. The average adjusted prediction (this is the stata-ese language - it is also called variously a predictive margin among other things). The adjusted prediction in general is just the predicted value of the dependent variable for some set of values on the covariates. In this case, I want to predict on the level of the response to get predicted probabilities of being held back. The average adjusted prediction essentially gets the adjusted prediction for every observation in the data based on their covariates and then takes the average. This is equivalent to treating every respondent as if they belonged to group A when calculating the average adjusted prediction for group A. The average adjusted prediction is directly related to the *average marginal effect* (AME) in that the AME for a group difference is simply the difference in the average adjusted prediction across groups.

@@

I have created a function that computes these average adjusted predictions, calculates midpoint values and SEs and does some other computaations described below. It spits out a list with a `coef_table` showing the average adjusted predictions and a `ci_df` which provides additional information needed for the graphs. Basically, this output will allow me to create the plots I need for each biracial comparison.

@@

Now I can use this output to plot out the results. I illustrate what this basic plot looks like below for the case of black/white individuals.

@@

This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions (although the standard errors are higher due to survey design effects of clustering and variable weights). The dots give the point estimates for all three groups. You can see from the graph that the point estimate of grade retention risk for Black/White students is much closer to the lower for White students than the higher risk for Black students.

The figure also provides confidence bands that help to visualize whether the difference between groups are statistically significant. People often believe that an overlap in the 95% confidence interval means the difference between the parameters is not statistically significant at the 5% level. This is actually a mistake and leads to far more conservative assessments. The actual p-value of rejection corresponding to an overlap of 95% CI is about 0.0055 when the SEs of the two estimates are the same as shown in [this paper](https://link.springer.com/article/10.1007/s10654-011-9563-8).

If the SEs of the two estimates are the same (and they are independent), then there is a simple correction to this problem: use 83.4% confidence interval (z=1.386). In that case, lack of overlap in CIs indicates that the differences between estimates are statistically significant at the 5% level. Overlap indicates that they are not statistically significant.

While this provides a nice and easy solution, it unfortunately only works in the case where the SEs are the same. Its a pretty good approximation when those SEs are close, but in my case, I get very different SEs due to massive differences in sample size across racial groups. To calculate the z-score for the confidence interval needed in the case of unequal SEs (and uncorrelated), first one needs to calculate $\rho=se_1/se_2$ which gives the ratio of the two SEs. It doesn't matter which serves as the numerator and denominator as the results are symmetric. The required z-score is then given by:

$$z=1.96*\frac{\sqrt{1+\rho^2}}{1+\rho}$$

The problem is that this z-score will be different for each pairwise comparison, and I want to make at least two comparisons in the figures - I want to compare the biracial group to each of its constituent monoracial groups. So, the way I handle this in the figure above is to calculate two confidence intervals for each biracial group. Each monoracial group gets a confidence band corresponding to that comparison and the biracial group gets two half-confidence intervals.All of these are color coded for the correct comparison and in the biracial case, they are only shown in the relevant direction. So in the case, above, the purple line doesn't come close to overlapping with the purple band for Black students and so this difference is clearly statistically distinguishable. The yellow line also does not overlap with the yellow band for whites, so this difference is also statistically distinguishable, although as you can see from the graph, that is just barely the case.

To check the graph, I can calculate AMEs for the Black/White group from the same model where I treat White and Black students, respectively, as the reference.

```{r test-marg}
marg(model0, "race", type="effects",
     at_var_interest=c("White","Black/White"))
marg(model0, "race", type="effects",
     at_var_interest=c("Black","Black/White"))
```

As you can see, the results are consistent here.

The graph above also shows a grey band for the halfway point. To avoid getting cluttery, I do not put a separate confidence line on the Black/White estimate to differentiate this case. In general that line will be about halway between the other two, so I think its ok to just eyeball that one. If it annoys reviewer, it can be added.

@@

```{r cond-means-model4, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. Non-overlap in confidence bands of corresponding color indicates statistically significant difference at the 5% level. Estimates are based on models that adjust for state, year, current grade, metropolitan status, gender, foreign born status, parent's foreign born status, English proficieny, parent's English proficiency, parent's education, family income, home ownership, and family structure."}
results_mod4 <- calculate_marg_means(model4)

results_mod0$coef_table %>%
  ggplot(aes(x=term, y=estimate, color=colors))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower,
                ymax=upper, fill=colors), color=NA, alpha=0.2)+
  geom_linerange(data=results_mod0$ci_df, 
                 position=position_dodge2(width = 0.05), size=1.2,
                 aes(ymin=lower, ymax=upper, color=colors))+
  geom_point(size=2)+
  coord_flip()+
  facet_wrap(~mrace, ncol=3, scales="free_y")+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=color_choices)+
  scale_fill_manual(values=color_choices)+
  labs(x=NULL,
       y="average adjusted predicted probability of being behind expected grade")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())
```

@@

```{r cond-means-model-compare, fig.height=12, warning=FALSE, fig.cap="Probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups. Non-overlap in confidence bands of corresponding color indicates statistically significant difference at the 5% level. Each panel is based on a different model that accounts for certain variables. Subsequent panels cumulatively include all prior terms."}
#calculate remaining marginal means
coef_table_mod1 <- calculate_marg_means(model1)
coef_table_mod2 <- calculate_marg_means(model2)

@@

-   Lower than both groups
-   Same as lower group
-   Lower half
-   Halfway between
-   Upper half
-   Same as higher
-   Higher than both groups

@@

|                   | Lower than both | Same as lower | Lower half | Halfway | Upper half | Same as higher | Higher than both |
|---------|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|:-------:|
| Black/White       |       X\*       |               |            |         |            |                |                  |
| White/Asian       |       X\*       |       X       |            |         |            |                |                  |
| White/Latino      |        X        |      X\*      |            |         |            |                |                  |
| Black/Latino      |        X        |      X\*      |     X      |    X    |            |                |                  |
| Indigenous/Asian  |        X        |      X\*      |     X      |         |            |                |                  |
| Indigenous/White  |                 |               |    X\*     |         |            |                |                  |
| Indigenous/Latino |                 |               |     X      |   X\*   |     X      |                |                  |
| Black Asian       |                 |       X       |     X      |   X\*   |     X      |       X        |                  |
| Black/Indigenous  |                 |       X       |     X      |   X\*   |     X      |       X        |                  |
| Latino/Asian      |                 |       X       |     X      |    X    |     X      |       X        |       x\*        |

@@

-   For White/Black students, grade retention is clearly in the lower-half between Whites and the midway point rather than below whites for high school students. Notably, this is also where the white/black grows the largest. So the advantage that these students have relative to everyone else only holds for elementary and middle school. Nonetheless, their outcomes are still much closer to White students than Black students.
-   For White/Indigenous students, the placement varies across panels. In elementary school, these students are indistinguishable from White students. They are perfectly in the middle in middle school, and then in the lower half in high school. So it seems that their lowest outcomes happen earlier and on average you end up with a "lower half" across all grades.
